Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-30 trunk/paper/drinking.tex

Eventually, a set of {\em primitive terms} remains that are not
further defined. Using only the primitive terms, we can construct
complex sentences. Based on the intended meaning of the primitive
terms, we consider some of these sentences true and some of them false
in our domain. We select some of the true sentences as {\em axioms}
which provide the core of our ontology. Ideally, the axioms are chosen
so that all true sentences in the domain we intend to represent follow
by means of logical deduction from the axioms. More commonly, however,
only {\em some aspects} of the intended meaning are formally
represented while other aspects are omitted either due to limitations
in language expressivity or due to their irrelevance to the problem
for which an ontology is developed.

Based on the axioms and definitions, we can use deduction to infer
statements that logically follow from the axioms.  The process of
automatically deducing sentences from axioms is called {\em automated
  reasoning}. Automated reasoning allows users of an ontology to carry
our key activities: verifying the ontology's consistency, inferring
hidden knowledge and thereby performing powerful queries.  An ontology
is inconsistent if {\em every} statement can be inferred from the
ontology. In particular, an ontology is inconsistent if a statement
$\phi$ and its negation $\neg \phi$ can be inferred from the
ontology's axioms. Classes in an ontology may be unsatisfiable: a
class $C$ is unsatisfiable, if it is not possible that the class has
instances. Unsatisfiable classes are commonly the result of a
contradictory class definition. That a class is unsatisfiable is often
a special kind of unintended consequence that can be drawn from an
ontology.

@@

Amongst abnormal physiology, we distinguish between abnormalities in a
{\em single occurrence} of a cell component's functioning and an
abnormal {\em pattern of multiple occurrences} of a cell component's
functioning. For example, abnormalities in cell division resulting in
{\em Aneuploidy} refer to abnormalities of functioning processes,
while an {\em increased rate of cell division} refers to an
abnormality in the pattern of occurrence of multiple cell division
processes. In the CPO, abnormalities in the pattern of occurrence of
$X$ are represented as abnormalities in the {\em regulation of $X$}.

>>>>

Eventually, a set of {\em primitive terms} remains that are not
further defined. Following the axiomatic method \cite{Hilbert1918},
using only the primitive terms, we can construct complex
sentences. Based on the intended meaning of the primitive terms, we
consider some of these sentences true and some of them false in our
domain. We select some of the true sentences as {\em axioms} which
provide the core of our ontology. Ideally, the axioms are chosen so
that all true sentences in the domain we intend to represent follow by
means of logical deduction from the axioms. More commonly, however,
only {\em some aspects} of the intended meaning are formally
represented while other aspects are omitted either due to limitations
in language expressivity or due to their irrelevance to the problem
for which an ontology is developed.

Based on the axioms and definitions, we can use deduction to infer
statements that logically follow from the axioms.  The process of
automatically deducing sentences from axioms is called {\em automated
  reasoning}. Automated reasoning allows users of an ontology to carry
our key activities: verifying the ontology's consistency, inferring
hidden knowledge and thereby performing powerful queries.  An ontology
is formally inconsistent if there is a statement $\phi$ such that
$\phi$ and its negation $\neg \phi$ can be inferred from the
ontology's axioms.  If an ontology is formally inconsistent, {\em
  every} statement can be inferred from the ontology.

Automated reasoning can further determine whether classes in an
ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
impossible for the class to have any instances. Unsatisfiable classes
in an ontology are commonly the result of a contradictory class
definition. % That a class is unsatisfiable is often a special kind of
% unintended consequence that can be drawn from an ontology.

@@

Amongst abnormal physiology, we distinguish between abnormalities in a
{\em single occurrence} of a cell component's functioning and an
abnormal {\em pattern of multiple occurrences} of a cell component's
functioning. For example, abnormalities in cell division resulting in
{\em Aneuploidy} refer to abnormalities of {\em cell division}
processes, while an {\em increased rate of cell division} refers to an
abnormality in the pattern of occurrence of multiple cell division
processes. In the CPO we follow the GO and represent abnormalities in
the pattern of occurrence of $X$ as abnormalities of {\em regulation
  of $X$} processes. In particular, an {\em increased rate of cell
  division} is not an attribute of {\em cell division} processes, but
rather an attribute of the {\em regulation of cell division}.
